---
title: "Calculate moderated nz-score"
output: html_notebook
---


jbagnall@broadinstitute.org, partially adapted from scripts by Marek Orzechowski</br>
November 27, 2023</br>
Calculate moderated nz-score for RNAseq data. This notebook is to be run after 01_qc_and_vst.Rmd. This notebook performs the following: <br />
<ol>
  <li>Robust nz-score</li>
  <li>Moderated z-score, collapsing replicates based on correlation between replicates</li></ol>

# User inputs
```{r}
#working directory
wkdir = '/idi/cgtb/jbagnall/psa_RNAseq/git/psa_rnaseq_manuscript_rproject/'

#output directory
outdir = '/broad/hptmp/jbagnall/Project/'

#run_name is something to identify this data set, will be included in saved file names
run_name = "Project"

#project_id0 could be the same as run_name, but is something corresponding to a field in the input data frame to identify this data set
project_id0 = "Project"

#vst file path, VST data calculated in 01_qc_and_vst.Rmd script
vst_filepath = paste0(outdir, paste(run_name, "vst_remove_low_count_samples_and_genes_n24x5894.gct", sep = "_"))

#Column name in the column metadata from the VST file that identifies the batches within which you will calculate z-scores, e.g. something that identifies the 384 well plates, or different data sets
batch_column_name = "project_id"

#Compound name (i.e. pert_id, perturbation ID) of negative control, to which the normalization will occur
negcon_name = "dmso"

#Value at which to clip the absolute value of the resulting z-scores, so outliers don't throw off down-stream correlations to the reference set
clip_value = 500
                     
```


# Set up working directory
```{r, setup, include=FALSE}
# Make sure working directory exists, otherwise create it
end_char = substr(wkdir, nchar(wkdir), nchar(wkdir))
if(end_char != "/"){
  wkdir = paste0(wkdir, '/')
}

if(dir.exists(wkdir)){
  setwd(wkdir)
  knitr::opts_knit$set(root.dir = wkdir) #This needs to be in the setup chunk to make it global for all chunks
}else{
  stop("Working directory does not exist")
}
print(paste0("Working directory: ", getwd()))
```



# Load libraries & data
```{r, include=FALSE}
library(cmapR)
library(dplyr)
library(ggplot2)
source('./functions/Functions_general.R')
source('./functions/Functions_mod_z_score.R')

#make sure output directory exists
end_char = substr(outdir, nchar(outdir), nchar(outdir))
if(end_char != "/"){
  outdir = paste0(outdir, '/')
}

if(!dir.exists(outdir)){
  print("Creating output directory")
  dir.create(outdir)
}

getwd()

#Load VST data calculated in script_qc_and_vst_before_moderated_zscore.Rmd
data_vst = parse_gctx(vst_filepath)
data_mat = data_vst@mat
row_meta = data_vst@rdesc
col_meta = data_vst@cdesc

```


## Group data by normalization units (comb_id)
```{r}
#Add a column "comb_id" that identifies the batch within which to normalize/calculate z-scores
temp = col_meta[[batch_column_name]]
col_meta[["comb_id"]] = temp
remove(temp)

#Assign batch to column "comb_id"
# length(unique(col_meta$comb_id))
# unique(col_meta$comb_id)

# col_meta = mutate(col_meta, comb_id = batch_column)
comb_id_unique = unique(col_meta$comb_id)
print(comb_id_unique)

#update metadata in gctx file
data_vst@cdesc = col_meta
```

## Calculate NZ-score: Z-score across each gene relative to DMSO/negative control 
### Prepare with QC
```{r}
#Each row is a gene, visualize metrics before using them
row_medians = rowMedians(data_mat, na.rm = T)
plot(x = seq(1:length(row_medians)), y = row_medians)

row_mads = apply(data_mat, MARGIN = 1, FUN = function(x){mad(x, na.rm = T)})
plot(x = seq(1:length(row_mads)), y = row_mads)

```

### Perform Z-score calculation per row
```{r}
#Subset to negative controls for input into the z-score function to get median and mad per row (gene)
col_meta_negcon = dplyr::filter(col_meta, pert_id == negcon_name)
unique(col_meta_negcon$pert_id)
View(table(col_meta_negcon$comb_id))
cidx = which(col_meta$id %in% col_meta_negcon$id)
data_mat_negcon = data_mat[,cidx]
 
zscore_param_list = list()
negcon_median_list = list()
negcon_mad_list = list()
idx_all_tracker = list()
data_zscore_negcon = data_mat_negcon #should I load this from file instead
data_zscore = data_mat

flag0 = rep(0, length(comb_id_unique)) #flag to check if all comb_ids ran
for(num in 1:length(comb_id_unique)){
    idx_negcon = which(col_meta_negcon$comb_id == comb_id_unique[num])
    idx_all = which(col_meta$comb_id == comb_id_unique[num])
    idx_all_tracker[[num]] = idx_all
    sum(length(idx_negcon))
    if(sum(length(idx_negcon))>0){
       zscore_param_list[[num]] = robust_zscore(data_mat_negcon[,idx_negcon], dim0 = 1) #z-score along rows, outputs zscores, medians and mads
       negcon_median_list[[num]] = zscore_param_list[[num]][[2]]
       negcon_mad_list[[num]] = zscore_param_list[[num]][[3]]
       data_zscore_negcon[,idx_negcon] = zscore_param_list[[num]][[1]]
       
       #Calculate z-scores
       data_zscore_sub = data_zscore[,idx_all]
       data_zscore_sub = sweep(data_zscore_sub, MARGIN = 1, STATS = negcon_median_list[[num]], FUN='-')
       data_zscore_sub = sweep(data_zscore_sub, MARGIN = 1, STATS = negcon_mad_list[[num]], FUN='/')
       data_zscore[, idx_all] = data_zscore_sub
       
    }else{
      flag0[num] = 1
    }
}

#make sure all the indices were covered once
idx_all_tracker = unlist(idx_all_tracker)
idx_all_tracker = sort(idx_all_tracker)
length(idx_all_tracker)
if(all(idx_all_tracker == seq(from = 1, to = length(idx_all_tracker), by = 1))){
  print("NZ-scores calculated for all indices")
}else{
  print("Warning: Some sample nz-scores not calculated")
}
if(any(duplicated(idx_all_tracker))){
  print("Warning: Some sample indices were repeated in nz-score calculation")
}

#Any comb_ids didn't get z-scored?
if(sum(flag0) > 0){
  print("Warning: Not all comb_ids were standardized")
}
```

#### Check nz-score distribution
```{r}
#Check distribution of nz-scores for QC, sanity checks
#Clip z-score values to account for extreme cases where genes with low count had low variation (i.e. low median absolute deviation)
#Look at genes' z-score distribution
#Smaller number of replicates leads to more spread z-score distribution, and may require higher clipping values.

#####In negcons only, check what are extreme zscore values#####
hist(data_zscore_negcon, breaks = 100)

#sample random gene nz-score distributions in the negative control samples
sample_gene_idx = sample(1:dim(data_zscore_negcon)[1], size = 10, replace = F)
for(n in sample_gene_idx){
  hist(data_zscore_negcon[n,], breaks = 100, main = paste(row_meta[n, 1], row_meta[n,3], row_meta[n,4], sep= ":"))
}
max(data_zscore_negcon)
min(data_zscore_negcon)

data_zscore_negcon_df = as.data.frame(data_zscore_negcon)
data_zscore_negcon_df$gene_id = rownames(data_zscore_negcon)
data_zscore_negcon_df = tidyr::gather(data_zscore_negcon_df, key = "original_id", value = "zscore", -gene_id)
data_zscore_negcon_df = left_join(data_zscore_negcon_df, col_meta, by = c("original_id" = "original_cid"))

gene_ids = unique(data_zscore_negcon_df$gene_id)
sample_gene_ids = sample(gene_ids, size = 10, replace = F)
for(n in sample_gene_ids){
  data_sub = dplyr::filter(data_zscore_negcon_df, gene_id == n)
  anno1 = dplyr::filter(row_meta, gene_id == n)
  p0 = ggplot(data_sub, aes(x = zscore))+
    geom_histogram()+
    # facet_wrap(~pert_id)+
    labs(title = paste(anno1$gene_id, anno1$gene_name, anno1$`gene_description_(refseq)`, sep = ":"))+
    theme_bw(base_size = 14)

  print(p0)  
}


#####in all samples######
sample_gene_idx = sample(1:dim(data_zscore)[1], size = 10, replace = F)
for(n in sample_gene_idx){
  hist(data_zscore[n,], breaks = 100, main = paste(row_meta[n, 1], row_meta[n,3], row_meta[n,4], sep= ":"))
}

#check range of zscores
max(data_zscore)
min(data_zscore)
iqr(data_zscore)
# quantile(data_zscore, c(0.005,0.01, .1, 0.2, 0.25, 0.5, 0.75, 0.8, 0.9, 0.99, 0.995)) #done later
median(data_zscore)


#how many zscores above clip_value (500)
data_zscore_df = as.data.frame(data_zscore)
data_zscore_df$gene_id = rownames(data_zscore_df)
data_zscore_df = tidyr::gather(data_zscore_df, key = "original_cid", value = "zscore", -gene_id)
data_zscore_df = left_join(data_zscore_df, col_meta, by = c("original_cid"))

high_zscore = filter(data_zscore_df, zscore > clip_value)
high_zscore_maxz = high_zscore %>%
  group_by(gene_id) %>%
  dplyr::slice(which.max(abs(zscore))) %>%
  ungroup() %>%
  arrange(desc(abs(zscore)))
length(unique(high_zscore$gene_id))
View(table(high_zscore$strain_id))
# 
# pdf(paste0(outdir, 'high_zscore_gene_histograms.pdf'), width = 8, height = 4)
# sample_gene_ids = high_zscore_maxz$gene_id
# for(n in sample_gene_ids){
#   data_sub = dplyr::filter(data_zscore_df, gene_id == n)
#   anno1 = dplyr::filter(row_meta, gene_id == n)
#   p0 = ggplot(data_sub, aes(x = zscore))+
#     geom_histogram(binwidth = 5)+
#     labs(title = paste(anno1$gene_id, anno1$gene_name, anno1$`gene_description_(refseq)`, sep = ":"), xlab = "Z-score to neg. control", ylab = "# Samples")+
#     theme_bw(base_size = 14)
# 
#   print(p0)  
# }
# dev.off()

#hist of all zscores
hist(data_zscore, breaks = 100)
hist(data_zscore, breaks = 100, ylim = c(0,100))

```


### Plot PCA of the nz-scores
```{r}
data_zscore_pca = run_pca(data_zscore, col_meta = col_meta, color_col = "pert_id", scale0 = T)
```

### Save z-scores into GCT file
```{r}
if(!all(colnames(data_zscore) == col_meta$id)){
  print("Warning: column names of matrix & metadata IDs do not match")
}
if(!all(rownames(data_zscore) == row_meta$gene_id)){
  print("Warning: row names do not match gene IDs in metadata")
}

save_to_gct(data_mat = data_zscore, data_cdesc = col_meta, data_rdesc = row_meta, savefilepath = paste0(outdir, paste(run_name, "zscore_to_negcon_remove_low_count_samples_and_genes", sep = "_")))

num_row_zscore = dim(data_zscore)[1]
num_col_zscore = dim(data_zscore)[2]

```


##Clip extreme z-scored values (should make sense with distribution visualized earlier, only affecting < 1% of z-scores)
```{r}
zscore = cmapR::parse_gctx(paste0(outdir, paste(run_name, "zscore_to_negcon_remove_low_count_samples_and_genes", sep = "_"), "_n", num_col_zscore, "x", num_row_zscore, ".gct"))

col_meta = zscore@cdesc
row_meta = zscore@rdesc
data_zscore = zscore@mat

quantile(data_zscore, c(0.001, 0.005, 0.01, 0.05, 0.95, 0.99, 0.995, 0.999))

#how many scores above 500? should be less than 1%
# clip_value = 500 #defined by user input in first chunk
# % of treatments affected by clipping
num_above = sum(data_zscore > clip_value)
num_below = sum(data_zscore < (-1*clip_value))
num_total = length(data_zscore)
percent_clipped = 100*(num_above + num_below)/num_total
print(paste0("percent clipped: ", signif(percent_clipped, digits = 4), "%"))

#Clip z-scores
data_zscore = base::pmax(data_zscore, (-1*clip_value))
data_zscore = base::pmin(data_zscore, clip_value)

save_to_gct(data_zscore, data_cdesc = col_meta, data_rdesc = row_meta, savefilepath = paste0(outdir, paste(run_name, 'zscore_to_negcon_remove_low_count_samples_and_genes_clipped', clip_value, sep = "_")))

```


## Collapse replicates
```{r}
zscore = cmapR::parse_gctx(paste0(outdir, paste(run_name, 'zscore_to_negcon_remove_low_count_samples_and_genes_clipped', clip_value, sep = "_"), "_n", num_col_zscore, "x", num_row_zscore, ".gct"))

col_meta = zscore@cdesc
data_zscore = zscore@mat
max(data_zscore)
min(data_zscore)

#Make sure replicates are identified properly
length(unique(col_meta$condition_id))
View(table(col_meta$condition_id))

#identify replicates
#perform moderated collapsing of averages
condition_id_unique = unique(col_meta$condition_id)
num_genes = dim(data_zscore)[1]
num_samples = dim(data_zscore)[2]
num_conditions = length(condition_id_unique)

data_rep_collapse = matrix(NA, nrow = num_genes, ncol = num_conditions)

all(col_meta$id == colnames(data_zscore))

for(num in 1:num_conditions){
  idx = which(col_meta$condition_id == condition_id_unique[num])
  cidx = which(colnames(data_zscore) %in% col_meta$id[idx])
  data_rep_collapse[,num] = modzs_collapse_avg(data = data_zscore[,cidx], ridx = seq(1, num_genes, by = 1))
}

colnames(data_rep_collapse) = condition_id_unique
rownames(data_rep_collapse) = rownames(data_zscore)

```


### Save replicate collapsed data as gct file
```{r}
#Revamp column metadata since replicates are collapsed
col_meta_collapse = dplyr::select(col_meta, condition_id, pert_id, pert_dose, pert_idose, pert_dose_unit, project_id, strain_id, comb_id) %>% distinct()

col_meta_collapse_sub = data.frame(condition_id = condition_id_unique)
col_meta_collapse_sub = left_join(col_meta_collapse_sub, col_meta_collapse, by = "condition_id")
if(!all(colnames(data_rep_collapse) == col_meta_collapse_sub$condition_id)){
  print("Warning: not all column names match metadata IDs in metadata")
}
if(!all(rownames(data_rep_collapse) == row_meta$gene_id)){
  print("Warning: not all row names match gene IDs in metadata")
}

save_to_gct(data_mat = data_rep_collapse, data_cdesc = col_meta_collapse_sub, data_rdesc = row_meta, savefilepath = paste0(outdir, paste(run_name, "modzscore_remove_low_count_samples_and_genes_rep_collapse_clipped", clip_value, sep = "_")))

num_row_collapse = dim(data_rep_collapse)[1]
num_col_collapse = dim(data_rep_collapse)[2]
```

### Plots for visualizing after replciates combined
```{r}
col_meta_collapse_sub$id = col_meta_collapse_sub$condition_id
data_zscore_rep_collapse_pca = run_pca(data_rep_collapse, col_meta = col_meta_collapse_sub, color_col = "pert_id", scale0 = T)
```

### Add inactive info and dose ranks to the column metadata if desired (optional)
```{r}
data_rep_collapse_path = paste0(outdir, paste(run_name, "modzscore_remove_low_count_samples_and_genes_rep_collapse_clipped", clip_value, sep = "_"), "_n", num_col_collapse, "x", num_row_collapse, ".gct")
data_rep_collapse = parse_gctx(data_rep_collapse_path)
data_rep_collapse_mat = data_rep_collapse@mat

#Determine weak signal separately based on DESeq2 p-values
inactives = readRDS('inactives.rds')
inactives_filt = filter(inactives, numerator_project_id == tolower(project_id0))

#condition_ids changed to include 5 digits after the decimal in the pert_dose to be consistent with previous runs
inactives$numerator_condition_id_old = inactives$numerator_condition_id
inactives = mutate(inactives, numerator_condition_id=paste(numerator_project_id, numerator_pert_id, paste0(sprintf("%.5f", round(numerator_pert_dose, 5)), "uM"), numerator_strain_id, paste0(numerator_pert_time, "min"), "A", sep = ":"))

cdesc1 = data_rep_collapse@cdesc
colnames(cdesc1)
colnames(inactives)
any(tolower(cdesc1$condition_id) %in% tolower(inactives$numerator_condition_id))
all(tolower(inactives_filt$numerator_condition_id) %in% tolower(cdesc1$condition_id))

cdesc1 = mutate(cdesc1, inactive = ifelse(tolower(condition_id) %in% tolower(inactives$numerator_condition_id), TRUE, FALSE))
sum(cdesc1$inactive)

cdesc1 = cdesc1 %>%
  group_by(strain_id, pert_id, pert_time, pert_itime) %>%
  arrange(pert_dose) %>%
  mutate(dose_rank = 1:n()) %>% 
  ungroup()

#By definition
# cdesc1 = mutate(cdesc1, inactive = ifelse(pert_type == "negcon", TRUE, inactive))
# sum(cdesc1$inactive)

any(cdesc1$id %in% colnames(data_rep_collapse@mat))
all(cdesc1$id %in% colnames(data_rep_collapse@mat))

#reorder cdesc
cidx = match(colnames(data_rep_collapse@mat), cdesc1$id)
cdesc1 = cdesc1[cidx, ]
all(cdesc1$id == colnames(data_rep_collapse@mat))
str(cdesc1)

data_rep_collapse@cdesc = as.data.frame(cdesc1)

write_gct(data_rep_collapse, paste0(outdir, paste(run_name, run_name_suffix, 'modzscore_remove_low_count_samples_and_genes_rep_collapse_inactivelabel_clipped', clip_value, sep = "_")))
```